---
title: "HW04: Tidy data and joins"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data reshaping

**Problem:** You have data in one “shape” but you wish it were in another. Usually this is because the alternative shape is superior for presenting a table, making a figure, or doing aggregation and statistical analysis.

**Solution:** Reshape your data. For simple reshaping, gather() and spread() from tidyr will suffice. Do the thing that is possible / easier now that your data has a new shape.

**Prompt:** Make a tibble with one row per year and columns for life expectancy for two or more countries from gapminder. 



First we load the tidyverse package and gapminder dataframe.
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

```{r}
fr_uk_lifeExp <- gapminder %>% 
  rename(Year = year) %>% 
  filter(country == "France" | country == "United Kingdom") %>% 
  select(Year, country, lifeExp) %>% 
  spread(key = "country", value = "lifeExp")
  colnames(fr_uk_lifeExp)
```

 - Use `knitr::kable()` to make this table look pretty in your rendered homework.
```{r}
knitr::kable(fr_uk_lifeExp)
```
 
- Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.
```{r}
fr_uk_lifeExp %>% 
ggplot(aes(`United Kingdom`, France)) +
  geom_point()
```

##Join Prompts

**Problem:** You have two data sources and you need info from both in one new data object.

**Solution:** Perform a join, which borrows terminology from the database world, specifically SQL.

Prompts:

Activity #1

Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:
        One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.
        One row per continent, a continent variable and one or more variables with extra info, such as northern versus southern hemisphere.
        
We shall build a second data frame of countries, with national animal, NATO membership


```{r}
#x1 <- unique(select(gapminder, continent))
#`Northern Hemisphere` <- c("Yes", "Yes", "Yes", "Yes", "No")
#`Southern Hemisphere` <- c("Yes", "No", "Yes", "Yes", "Yes")
#mutate(x1,`Northern Hemisphere`, `Southern Hemisphere`)

#Creating our own tibble

#column of chosen countries
countries <- c("Canada", "United Kingdom", "France", "Russia")

#column of continents corresponding to chosen countries
continent<- c("Americas", "Europe", "Europe", "Asia") 

#Lan <- c("English, French", "English", "French") `Official Language(s)` = Lan

#column of national animal of corresponding country
Anim <- c("Beaver", "Lion", "Gallic Rooster", "Bear") #national animal of corresponding coutries chosen above

#column of NATO membership
NATO <- c("Yes", "Yes", "Yes", "No") 

#column of Capital city
Cap <- c("Ottawa", "London", "Paris", "Moscow") 

#create the tibble `addinfo` consisting of all the above columns
(addinfo <- tibble(continent = continent, country = countries,  `national animal` = Anim, `NATO membership` = NATO, `capital city` = Cap))
```
```{r}
#filter gapminder dataset to year 2007, and the countries in the countries column created earlier
(gapfilt <- gapminder %>% 
  filter(year == 2007, country%in%countries | country == "Norway") %>% 
  
  #select subset of data in 2007
  select(continent, country, lifeExp))
```


```{r}
#left join by country
left_join(gapfilt, addinfo, by = "country")
```
```{r}
left_join(gapfilt, addinfo)
```
```{r}
left_join(addinfo, gapfilt)
```



```{r}


#left_join(addinfo, gapfilt , by = "country")


right_join(gapfilt, addinfo)
```

```{r}
right_join(addinfo, gapfilt)
```

```{r}
right_join(gapfilt, addinfo)
```

```{r}
inner_join(gapfilt, addinfo)
```


```{r}
inner_join(addinfo, gapfilt)
```

```{r}
full_join(gapfilt, addinfo)
bind_rows(gapfilt, addinfo)
```


```{r}
full_join(addinfo, gapfilt)
```


```{r}
semi_join(gapfilt, addinfo)
```

```{r}
semi_join(addinfo, gapfilt)
```


```{r}
anti_join(gapfilt, addinfo)
```

```{r}
anti_join(addinfo, gapfilt)
```

```{r}
bind_cols(gapfilt, addinfo)
```



```{r}
(a <- select(gapfilt, continent, country))
(b <- select(addinfo, continent, country))
```

```{r}
intersect(a,b)
```

```{r}
union(a,b)
```

```{r}
setdiff(a,b)
```

```{r}
setdiff(b,a)
```

#intersect(addinfo, gapfilt)
#union()
#setdiff()




